Deriving an Executable Model of a Data Stream Protocol from its Specification

### 2009-10-06 papers/ModelBasedTestCaseGenerationDataStreamProtocol/paper.tex
ModelBasedTestCaseGenerationDataStreamProtocol paper: discussion on hand made and generated tests

The use of both test suites, hand made and generated, has proven 
effective on finding protocols errors on the \STSB when the board 
prototypes have been delivered.

@@

The second prototype delivered was meant to build packets with values
taken from rear sensors installed on the board.  The only sensors that
were not yet working were the sensors devoted to the fast rate data
streams, that is, the audio sensors.  Only one error needed to be
corrected in the driver implementation, because of a (rare)
combinations of conditions that were not covered by the test cases.
One error have been discovered in the low level driver that the driver
was using; to eliminate the error an upgrade of the low level drover
to a new unreleased version was necessary.  No other errors were
detected on the driver side. All the errors on the \SB side have been
detected.

@@

\todo{describe why such a difference: random selection, many tests 
pruned if pre condition is not satisfied}

\todo{compare hand made, many calls, vs generated, single call, tests}

\todo{describe measures: coverage, number of new bugs found}

@@

The test effectiveness evaluation criteria is of two types.

\todo{describe quantitative: coverage}

\todo{describe qualitative: errors found, help in design, ...}

>>>>

JML specifications, the simulators,  and both test suites has been used to:

\begin{itemize}
\item verify hardware implementation;
\item verify driver implementation, with and without the physical
  board;
\item guide driver development.
\end{itemize}

@@

The second prototype delivered was meant to build packets with values
taken from real sensors installed on the board.  The only sensors that
were not yet working were the sensors devoted to the fast rate data
streams, that is, the audio sensors.  Only one error needed to be
corrected in the driver implementation, because of a (rare)
combinations of conditions that were not covered by the test cases.
One error have been discovered in the low level system driver that the
developed driver was using (the system driver was outside of our
development scope); to eliminate the error an upgrade of the low level
driver to a new unreleased version was necessary.  No other errors
were detected on the driver side. All the errors on the \SB side have
been detected.

@@

The ratio of 1 to 100, that is, the big number of generated tests, can
be understood looking closely at how the generated tests are created
by the JML framework. The generated tests are exploring all the
possible input combinations on public methods, combining the type data
ranges that have been specified for the test suite (see
\cite{Cheon-Leavens02} for more details on how the JML framework works
on generating unit tests). For instance, let's suppose we want to test
a method \lil{m(int p1, int p2)} for a class \lil{C}; the data ranges
specified for type \lil{int} are \lil{\{1,2,3\}} and for type \lil{C}
are the instances \lil{\{c1,c2,c3,c4\}}. The JML framework will
generate $24 = 3 \times 3 \times 4$ tests, the combination of the data
ranges involved (the data range of the parameters and the data range
of the type owning the method under test). But not all the generated
tests are executed.

A generated test, as it was previously mentioned (see section
\ref{sec:test_cases}), is not executed, unless the preconditions of
the called method are met. Thus, depending on the preconditions
specified for a method, the number of the generated tests that are
actually executed in the test suite can be significantly reduced.

To compare in a quantitative way the effectiveness of the test suites,
code coverage metrics have been used; this is not the only measure
that have been considered, since it is almost unanimously accepted
that code coverage alone cannot asses the quality of a test suite
\todo{reference to: ``How to misuse code coverage'', ??? ``Coverage
  Metrics for Formal Verification''}.

Various coverage metrics exist, we analyzed the test suites with
statement code coverage (one of the simplest types)\footnote{The tool
  used to obtain statement code coverage metrics is
  \myhref{http://emma.sourceforge.net/}{Emma}.}: statement coverage
reports whether each executable statement has been encountered. This
code coverage metric is easy to implement, but lacks the capability to
explore the control flow graph associated to the code, that is,
decision branches are not explored. The statement code coverage
results are reported in the Table \ref{tab:statement_code_coverage}.

\begin{table}[htbp]
  \caption{Statement code coverage result, categorized by source package.}
  \label{tab:statement_code_coverage}
  \begin{center}
    \begin{tabular}{|c|c|c|c|}\hline
      & \textbf{\textit{hand}} & \textbf{\textit{generated}} &
      \textbf{\textit{total}} \\\hline
      \textbf{\textit{Driver}} & $15.9 \%$ & $6.3 \%$ & $15.9 \%$ \\\hline
      \textbf{\textit{StreamDriver}} & $76 \%$ & $79.7 \%$ & $88.5 \%$ \\\hline
      \textbf{\textit{SimulatorDriver}} & $64.9 \%$ & $44.2 \%$ &
      $71.2 \%$ \\\hline
      \textbf{\textit{Utility}} & $98.1 \%$ & $74.5 \%$ & $98.1 \%$ \\\hline
    \end{tabular}
  \end{center}
\end{table}

The quantitative and qualitative specification and test effectiveness
measures to verify system under development obtained
in the previous section need to be analyzed to be fully understood.

The coverage measures reported in Table
\ref{tab:statement_code_coverage} are not high, in absolute
values. This is because of two different effects: non public methods,
and interfaces. The effect of non public methods (to be precise:
package Java visibility) is that these methods are taken into account
as well as public and protected ones during code coverage calculation;
but these methods are not part of the interface, they are not meant to
be called by the system, and they are only used in object
initialization or during explicit object setup during unit tests. In
fact there are package and private code blocks that are not reachable
by construction from public and protected methods, and this is a
precise characteristic reflecting the code architecture, meant to ease
class instances setup and configuration for testing purposes. The used
statement coverage tool cannot be configured to distinguish these code
blocks, so this effect is unavoidable. The effect is mainly seen in
low \lil{SimulatorDriver} code coverage figures: the
\lil{SimulatorDriver} package has plenty of setup and initialization
non public methods.

Regarding interfaces: an interface contains no statements, so when an
interface method is called, it is the method of the implementation
class used that is actually covered. This effect is visible in
\lil{Driver} coverage figures, \lil{Driver} package contains
interfaces (that are not counted at all) and some very simple real
class (exceptions) that are not throughly tested: the net result is
extremely low statement code coverage figures because the very simple
classes are the only ones that are actually covered. This effect could
be easily avoided writing specific hand made tests for the simple
classes and including them in the classes for which tests are
generated.

The statement code coverage figures shows that the test suites do a
good job on the most important package, that is, the
\lil{StreamDriver}. The generated test suit reaching almost $ 80 \% $,
while the hand made tests have only a slightly lower coverage ratio: $
76 \% $. But the combination of test suites is interesting,
raising the coverage ratio to $ 88.5 \% $. This is a clear indication
that one test suite is not completely overlapping the other; that is,
the test suites are not to be considered alternatives, but they have
to be used together to achive the higher benefits, they are
complementary. This consideration can be confirmed looking at the
figures obtained for \lil{SimulatorDriver} package as well.

The qualitative difference of the two test suites can be understood
analysing how the suites are built. The hand made test suite primarly
focuses on small features and use cases fragments that the system must
implement, or for which an exceptional case has been considered. A
test case is to be considered a rough specification written in
imperative style. The common structure require an object
initialization, followed by a hopefully not too long sequence of
public methods to bring one object into a desired state, than the
method under test is called and the object state is finally inspected
to verify it is in the expected state. The generated tests have a far
more limited scope, since only one method call is performed. Thus it
is difficult to explore complex behaviour. It is true that with a
complete specification, and specifing large enough data ranges types,
the genrated tests could teoretically achive the same expressive power
as the hand made ones, but the price to pay would be expensive: much
longer an more complex specifications, requiring more time to be
designed and or updated, and more tests to be run, augmentig the
already not so short computation time of the generated test suite.

\todo{compare qualitative: hand made, many calls, vs generated, single call, tests}

@@

\todo{comment quantitative: number of bugs found}

\todo{compare qualitative: spec driven dev, test driven dev, test
  oriented code is an automatic result}

\todo{spec and hand useful to drive, spec and hand and simulator
  useful to test, spec and generated useful for statement coverage and
  easy methods, spec useful to test hardware, hand useful to spec,
  spec useful to hand... test driven spec}
